\begin{thebibliography}{12}
\newcommand{\enquote}[1]{``#1''}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\providecommand{\urlprefix}{URL }
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi
\providecommand{\eprint}[2][]{\url{#2}}

\bibitem[{Allaire \emph{et~al.}(2018)Allaire, Eddelbuettel, and
  Fran\c{c}ois}]{CRAN:Rcpp:Attributes}
Allaire JJ, Eddelbuettel D, Fran\c{c}ois R (2018).
\newblock \emph{{Rcpp} Attributes}.
\newblock Vignette included in R package Rcpp,
  \urlprefix\url{http://CRAN.R-Project.org/package=Rcpp}.

\bibitem[{Bates and Eddelbuettel(2013)}]{JSS:RcppEigen}
Bates D, Eddelbuettel D (2013).
\newblock \enquote{Fast and Elegant Numerical Linear Algebra Using the
  {RcppEigen} Package.}
\newblock \emph{Journal of Statistical Software}, \textbf{52}(5), 1--24.
\newblock \urlprefix\url{http://www.jstatsoft.org/v52/i05/}.

\bibitem[{Bates \emph{et~al.}(2016)Bates, Fran\c{c}ois, and
  Eddelbuettel}]{CRAN:RcppEigen}
Bates D, Fran\c{c}ois R, Eddelbuettel D (2016).
\newblock \emph{RcppEigen: Rcpp integration for the Eigen templated linear
  algebra library}.
\newblock {R} package version 0.3.2.9.0,
  \urlprefix\url{http://CRAN.R-Project.org/package=RcppEigen}.

\bibitem[{Eddelbuettel(2013)}]{Eddelbuettel:2013:Rcpp}
Eddelbuettel D (2013).
\newblock \emph{Seamless R and C++ Integration with Rcpp}.
\newblock Use R! Springer, New York.
\newblock ISBN 978-1-4614-6867-7.

\bibitem[{Eddelbuettel and Balamuta(2018)}]{TAS:Rcpp}
Eddelbuettel D, Balamuta JJ (2018).
\newblock \enquote{Extending R with C++: A Brief Introduction to Rcpp.}
\newblock \emph{The American Statistician}, \textbf{72:1}.
\newblock \url{10.1080/00031305.2017.1375990}.

\bibitem[{Eddelbuettel and Fran\c{c}ois(2011)}]{JSS:Rcpp}
Eddelbuettel D, Fran\c{c}ois R (2011).
\newblock \enquote{{Rcpp}: Seamless {R} and {C++} Integration.}
\newblock \emph{Journal of Statistical Software}, \textbf{40}(8), 1--18.
\newblock \urlprefix\url{http://www.jstatsoft.org/v40/i08/}.

\bibitem[{Eddelbuettel \emph{et~al.}(2018)Eddelbuettel, Fran\c{c}ois, Allaire,
  Ushey, Kou, Russel, Chambers, and Bates}]{CRAN:Rcpp}
Eddelbuettel D, Fran\c{c}ois R, Allaire J, Ushey K, Kou Q, Russel N, Chambers
  J, Bates D (2018).
\newblock \emph{{Rcpp}: Seamless {R} and {C++} Integration}.
\newblock R package version 0.12.12,
  \urlprefix\url{http://CRAN.R-Project.org/package=Rcpp}.

\bibitem[{Eddelbuettel \emph{et~al.}(2016)Eddelbuettel, Fran\c{c}ois, and
  Bates}]{CRAN:RcppArmadillo}
Eddelbuettel D, Fran\c{c}ois R, Bates D (2016).
\newblock \emph{RcppArmadillo: Rcpp integration for Armadillo templated linear
  algebra library}.
\newblock R package version 0.7.400.2.0,
  \urlprefix\url{http://CRAN.R-Project.org/package=RcppArmadillo}.

\bibitem[{Eddelbuettel and
  Sanderson(2014)}]{Eddelbuettel+Sanderson:2013:RcppArmadillo}
Eddelbuettel D, Sanderson C (2014).
\newblock \enquote{{RcppArmadillo}: Accelerating {R} with High-Performance
  {C++} Linear Algebra.}
\newblock \emph{Computational Statistics and Data Analysis}, \textbf{71},
  1054--1063.
\newblock \doi{10.1016/j.csda.2013.02.005}.
\newblock \urlprefix\url{http://dx.doi.org/10.1016/j.csda.2013.02.005}.

\bibitem[{Galassi \emph{et~al.}(2010)Galassi, Davies, Theiler, Gough, Jungman,
  Alken, Booth, and Rossi}]{GSL}
Galassi M, Davies J, Theiler J, Gough B, Jungman G, Alken P, Booth M, Rossi F
  (2010).
\newblock \emph{{GNU} {S}cientific {L}ibrary {R}eference {M}anual}, 3rd
  edition.
\newblock Version 1.14. {ISBN} 0954612078,
  \urlprefix\url{http://www.gnu.org/software/gsl}.

\bibitem[{{R Core Team}(2018)}]{R:Main}
{R Core Team} (2018).
\newblock \emph{R: A Language and Environment for Statistical Computing}.
\newblock R Foundation for Statistical Computing, Vienna, Austria.
\newblock \urlprefix\url{https://www.R-project.org/}.

\bibitem[{Sanderson(2010)}]{Sanderson:2010:Armadillo}
Sanderson C (2010).
\newblock \enquote{{Armadillo}: {An} open source {C++} Algebra Library for Fast
  Prototyping and Computationally Intensive Experiments.}
\newblock \emph{Technical report}, {NICTA}.
\newblock \urlprefix\url{http://arma.sf.net}.

\bibitem[{Sklyar \emph{et~al.}(2015)Sklyar, Murdoch, Smith, Eddelbuettel, and
  Fran\c{c}ois}]{CRAN:inline}
Sklyar O, Murdoch D, Smith M, Eddelbuettel D, Fran\c{c}ois R (2015).
\newblock \emph{inline: Inline C, C++, Fortran function calls from R}.
\newblock R package version 0.3.14,
  \urlprefix\url{http://CRAN.R-Project.org/package=inline}.

\end{thebibliography}
